Hypoxia uncouples HIF gene transcription and metabolic flux in proliferating primary cells

### 2021-04-02 inst/analysis/flux-differences.Rmd
work on mid figures for manuscript

---
title: "Mass Isotopomer Distributions"
subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: true
    toc_depth: 1
    df_print: kable
    dev: cairo_pdf
    latex_engine: xelatex
geometry: margin = 0.5in
fontsize: 11pt
mainfont: Calibri
---

@@

```{r}
withr::with_dir(here::here(), {
  mids <- tar_read(mids)
})
```

@@

Mass isotopomer distributions (MIDs) were determined for key intracellular metabolites. Cells were seeded on Day -1 in "light" labeling medium. On Day 0, medium was changed to "heavy" labeling medium containing an equimolar concentration of the stable isotope substrate. These experiments utilized the following substrates: [1,2-^13^C~2~] glucose, [U-^13^C~6~] glucose, [U-^13^C~5~] glutamine, and [U-^13^C~3~] lactate. Intracellular metabolites for every time point were extracted in 80% MeOH pre-cooled to -80 °C and analyzed by LC-MS. Each sample was run using two different LC-MS methods. First, selected ion monitoring (SIM) was used to enhance the sensitivity for key metabolites used in modeling. Second, a polarity-switching, full scan (FS) method was used. The full scan method provided labeling insights for metabolites that were not previously targeted.

# Data Processing

Peaks were interpolated using either Thermo Xcalibur or TraceFinder software.

## Quadrupole bias correction

A selected ion monitoring method was utilized to increase sensitivity for isotope detection. Others have noted quadrupole bias on the Q Exactive using this detection method (*i.e.*, the sensitivity for each isotope differs relative to its position within the m/*z* window). To account for this, naturally labeled standards were analyzed using methods where the m/*z* window is shifted -1.003355 units, the ^13^C mass defect. The M1/M0 ratio was calculated for each frame shift and corrected to the theoretical ratio. These correction factors are applied to the raw MID data.

## Natural abundance correction matrices

Isotope correction matrices are generated from the metabolite's molecular formula using a custom R package (`wmoldham/mzrtools`).

## Correct for natural isotope abundance

MIDs are calculated from the peak areas corrected for quadrupole bias for each metabolite within each run. The correction matrices are are then used to adjust experiment MIDs for the natural abundance of ^13^C.

@@

```{r}
mids_filtered <-
  mids %>% 
  group_by(across(-c(date, mid))) %>% 
  remove_nested_outliers(mid, remove = TRUE)
```

```{r}
mids_filtered %>% 
  pivot_wider(names_from = method, values_from = mid) %>% 
  filter(!is.na(sim)) %>%  
  ggplot() + 
  aes(
    x = sim, 
    y = fs, 
    color = isotope
  ) +
  facet_wrap(~metabolite) +
  geom_abline(slope = 1) +
  geom_point() + 
  scale_color_manual(values = viridis::viridis(7)) +
  labs(
    x = "Selected ion monitoring", 
    y = "Full scan", 
    title = "Method comparison"
  )
```

There is reasonably good agreement between full scan and SIM methods for several of the metabolites. This observation suggests that those metabolites detected in full scan mode only are reasonable approximations. Certainly, SIM mode is preferred for metabolites detected with it, particularly as the MID correction seems to be better.

```{r remove-excess-fs}
fs <-
  mids_filtered %>% 
  filter(method == "fs") %>% 
  pull(metabolite) %>% 
  unique()

sim <- 
  mids_filtered %>% 
  filter(method == "sim") %>% 
  pull(metabolite) %>% 
  unique()

fs <- setdiff(fs, sim)

method_filter <-
  tibble(
    metabolite = c(fs, sim), 
    method = c(rep("fs", length(fs)), rep("sim", length(sim)))
  )

mids_method <-
  mids_filtered %>% 
  semi_join(method_filter, by = c("metabolite", "method"))
```

```{r tracer-label-key}
tracer_labels <- c(
  expression(paste("[1,2-"^13, "C"[2], "] glucose")),
  expression(paste("[U-"^13, "C"[6], "] glucose")), 
  expression(paste("[U-"^13, "C"[5], "] glutamine")), 
  expression(paste("[U-"^13, "C"[3], "] lactate"))
)
tracer_levels <- c("glc2", "glc6", "q5", "lac3")
```

```{r, results = "hide", fig.show = "hold", out.width = "49%"}
mids_method %>% 
  arrange(metabolite) %>% 
  filter((cell_type == "lf" & time == 72) | (cell_type == "pasmc" & time == 48)) %>% 
  group_by(method, cell_type, metabolite) %>% 
  mutate(tracer = factor(tracer, levels = tracer_levels, labels = tracer_labels)) %>% 
  nest() %>% 
  mutate(plots = pmap(
    list(data, metabolite, method, cell_type),   
    ~ ggplot(data = ..1) + 
      aes(
        x = isotope, 
        y = mid, 
        color = interaction(oxygen, treatment, sep = " | ")
      ) +
      facet_wrap(
        ~tracer, 
        nrow = 2, 
        ncol = 2, 
        scales = "free_x", 
        labeller = label_parsed
      ) +
      stat_summary(
        geom = "pointrange", 
        position = position_dodge(width = 0.3), 
        fun.data = "mean_se"
      ) +
      labs(
        x = "Isotope", 
        y = "Mole fraction", 
        title = stringr::str_c(toupper(..3), " ", ..2, " ", toupper(..4)), 
        color = NULL
      )
  )) %>% 
    pull(plots)
```

>>>>

---
title: "Model Analysis"
subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: true
    toc_depth: 1
    df_print: kable
    dev: cairo_pdf
    latex_engine: xelatex
geometry: margin = 0.5in
fontsize: 11pt
mainfont: Calibri
---

@@

```{r}
withr::with_dir(here::here(), {
  model_fluxes <- tar_read(model_fluxes)
})
```

@@

In this document, we will analyze the output from the metabolic flux maps.

```{r}
usethis::use_data(model_fluxes, overwrite = TRUE)
```

@@

## Significant Differences

Identify reactions that are significantly different based on whether the confidence intervals overlap. This is more conservative than p < 0.05.

### 2021-04-02 inst/manuscript/manuscript.Rmd
work on mid figures for manuscript

To test this hypothesis, we developed models of bioenergetic carbon flux in human primary cells cultured under 21% or 0.5% oxygen conditions. We found that hypoxia fails to increase glycolysis in primary cells despite robust up-regulation of the HIF-1α transcriptional program. In normoxia, HIF-1α activation by the prolyl hydroxylase inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response. These findings suggest the existence of hypoxia-dependent metabolic regulatory mechanisms that override the effects of HIF-1α-dependent up-regulation of glycolytic gene expression.

@@

## Extracellular flux analysis reveals little effect of hypoxia on glycolysis

Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (Figure \ref{fig:f1}I-J and Supplementary Figure 1). Flux calculations incorporated the changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (Supplementary Figure 1). Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (Figure \ref{fig:f1}I) despite activation of the HIF-1 transcriptional program. Similar findings were observed in pulmonary artery smooth muscle cells (Supplementary Figure 2) and when the ambient oxygen level was decreased further to 0.2% (Supplementary Figure 3). In addition to glucose and lactate, extracellular fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake. Notably, hypoxia was previously shown to increase glutamine uptake in studies of cancer cell metabolism [@RN1447; @RN2395; @RN628].

Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated changes in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in lung fibroblasts. Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (Figure \ref{fig:f2}). Similar to hypoxia, BAY decreased cell growth rate (Figure \ref{fig:f2}A-B) and activated the HIF-1 transcriptional program (Figure \ref{fig:f2}C-G). Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression. In normoxia, this transcriptional program was associated with increased glucose uptake and lactate efflux (Figure \ref{fig:f2}H). Relatively modest effects on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (Figure \ref{fig:f2}J).

>>>>

To test this hypothesis, we developed models of bioenergetic carbon flux in human primary cells cultured under 21% or 0.5% oxygen conditions. We found that hypoxia fails to increase glycolysis in primary cells despite robust up-regulation of the HIF-1α transcriptional program. In normoxia, HIF-1α activation by the prolyl hydroxylase inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response. These findings suggest the existence of hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1α-dependent up-regulation of glycolytic gene expression in human primary cells.

@@

## Extracellular flux analysis reveals little impact of hypoxia on glycolysis

Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (Figure \ref{fig:f1}I-J, Supplementary Figures 1-3). Flux calculations incorporated the changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (Supplementary Figure 1). Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (Figure \ref{fig:f1}I) despite activation of the HIF-1 transcriptional program. Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (Supplementary Figure 2) and in pulmonary artery smooth muscle cells (PASMCs; Supplementary Figure 3). In addition to glucose and lactate, the absolute extracellular fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake in lung fibroblasts. Notably, hypoxia was previously shown to increase glutamine uptake in studies of cancer cell metabolism [@RN1447; @RN2395; @RN628]. Glutamine uptake did increase in PASMCs, as did the uptake of branch chain amino acids and arginine, suggesting differential responses of these mesenchymal cells to hypoxia.

Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in lung fibroblasts. Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (Figure \ref{fig:f2}). Similar to hypoxia, BAY decreased cell growth rate (Figure \ref{fig:f2}A-B) and activated the HIF-1 transcriptional program (Figure \ref{fig:f2}C-G). Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression. In normoxia, this transcriptional program was associated with increased glucose uptake and lactate efflux (Figure \ref{fig:f2}H). Relatively modest effects on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (Figure \ref{fig:f2}J).

## Stable isotope tracing

To characterize further the extent of hypoxia-induced metabolic reprogramming, we next treated lung fibroblasts with stable carbon isotopes of glucose and glutamine to trace label incorporation into key carbon utilization pathways (Figure \ref{fig:f3} and Supplementary Figure 4). Overall, hypoxia-treated cells had decreased label incorporation into downstream metabolites (*i.e.*, the unlabeled, or M0, fraction was greater). This finding is consistent with the extracellular flux measurements suggesting slower substrate utilization by hypoxic cells. BAY treatment recapitulated the labeling pattern observed with hypoxia, suggesting similar intracellular metabolite flux between these two conditions. Beyond this observation, the labeling patterns in hypoxia- and BAY-treated cells were similar to their respective controls, arguing against marked metabolic reprogramming in response to prolyl hydroxylase inhibition by either hypoxia or BAY. Compared to previous studies of metabolic flux in cancer cells [@RN2395; @RN1447], no hypoxia-mediated increase in M5-labeled citrate from [U-^13^C~5~] glutamine was observed, indicating no increase in reductive carboxylation for lipid synthesis. The overall fraction of M5-citrate in these cells was low (< 6%).
